---
title: "STA286 Lecture 02"
author: "Neil Montgomery"
date: "Last edited: `r format(Sys.time(), '%Y-%m-%d %H:%M')`"
output: 
  beamer_presentation:
    incremental: TRUE
    df_print: tibble
    fig_caption: FALSE
classoption: aspectratio=169

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
options(tibble.width=70)
```

```{r, message=FALSE}
library(tidyverse)
library(readxl)

# Import the data
oil <- read_excel("oil_readings.xlsx")

# Fix up some variable formatting issues. A necessary annoyance!
oil <- oil %>% 
  mutate(Ident = factor(Ident), Date=as.Date(Date), TakenBy = factor(TakenBy))

# A long version of the data
oil_long <- oil %>% 
  gather(element, ppm, -Ident:-TakenBy) %>% 
  mutate(element = factor(element))
```

# numerical summaries of dataset variables --- definitions first with examples after

## sample measures of "location"

The dataset is often called the "sample" (no matter where the data came from).

\pause For a particular numerical variable in the sample with observations:
$$\{x_1,x_2,\ldots,x_n\}$$
the *sample average* is just the arithmetic mean:
$$\overline{x}=\frac{1}{n}\sum\limits_{i=1}^n x_i$$
\pause Could be sensitive to extreme observations. 

## sample medians, sample percentiles

Order the observations:

$$x_{(1)} \le x_{(2)} \le \cdots \le x_{(n)}$$
A number that divides the observations into two groups is called a *sample median*. For example:
$$\tilde{x} = \begin{cases}
x_{\left((n+1)/2\right)} &: n \text{ odd}\\
\left(x_{\left(n/2\right)} + x_{\left(n/2 + 1\right)}\right)/2 &: n \text{ even}
\end{cases},$$
which is harder to write out than it is to understand.

\pause A *sample $p^{th}$ percentile* has $p$\% of the data below or equal to it. Special cases include (sample...): quartiles, quintiles, deciles, and indeed the median itself.

## sample measures of variation of a numerical variable

Very (too?) simple measure: *sample range* which is just $x_{(n)} - x_{(1)}$.

\pause More common to consider the set of deviations from the sample mean:
$$x_i - \overline{x}$$
Adding them up just gives 0, so instead consider positive functions such as:
$$|x_i - \overline{x}|\qquad \text{ or } \qquad (x_i - \overline{x})^2$$
\pause Summing up over all the observations gives the *sum of absolute deviations* (aka SAD) and the *sample variance* respectively. Notation and formula:
$$s^2 = \frac{\sum\limits_{i=1}^{n} \left(x_i - \overline{x}\right)^2}{n-1}$$

## sample standard deviation

$s^2$ is essentially the average squared deviation. (More on $n-1$ later in the course.)

The sample variance is good for theory but has an inconvenient unit. More practical is the *sample standard deviation*:
$$s = \sqrt{s^2}$$

## numerical summaries for categorical variables

The oil readings data had one categorical variable, the `Ident` variable which is just a serial number. I added a fake one `TakenBy` for illustration.

```{r}
oil[1:5,]
```

## tables of counts (or proportions)

A categorical variable could also be called a *factor* variable with *levels*, and to tabulate the frequency of each level is the way to summarize.

```{r}
oil %>% 
  count(Ident) %>% 
  mutate(proportion = n/sum(n))
```

## two-way classification with `Ident` and `TakenBy`

```{r}
with(oil, table(TakenBy, Ident))
```

# graphical summaries

## barchart

A barchart is a table of counts, in graphical form.

```{r, fig.height=4}
oil %>% 
  ggplot(aes(x=Ident)) + 
  geom_bar() + 
  theme(axis.text.x = element_text(angle=90))
```

## "Pareto" chart 

Ordered by count.

```{r}
oil %>% 
  ggplot(aes(x=reorder(Ident, -table(Ident)[Ident]))) + 
  geom_bar() + 
  theme(axis.text.x = element_text(angle=90)) + 
  xlab("Ident")
```

## piecharts are problematic

```{r, fig.height=8, fig.asp=1}
pie(table(oil$Ident))
```

## histograms

A histogram is a special case of a barchart.

A numerical variable is split into classes and a barchart is made from the table of counts of obvservations within each class. 

Histograms are done by the computer. Always play around with the number of classes.

```{r, fig.width=3.8, fig.asp = 0.618034}
oil %>% 
  ggplot(aes(x=Fe)) + geom_histogram(bins=50, color="black", fill="white") 
```

## histograms are hard to implement!

Better picture around 0. Possibly not important for EDA?

```{r}
oil %>% 
  ggplot(aes(x=Fe)) + geom_histogram(bins=50, boundary=0, colour="black", fill="white")
```

## histogram without those really big values

```{r}
oil %>% 
  filter(Fe < 100) %>% 
  ggplot(aes(x=Fe)) + geom_histogram(bins=50, boundary=0)
```

## a few more ppm histograms

```{r}
## Special function for more than one plot on the same page.
source("multiplot.R")

p1 <- oil %>% 
  ggplot(aes(x=Si)) + geom_histogram(bins=50)

p2 <- oil %>% 
  ggplot(aes(x=Ca)) + geom_histogram(bins=50)

p3 <- oil %>% 
  ggplot(aes(x=Cu)) + geom_histogram(bins=50)

p4 <- oil %>% 
  ggplot(aes(x=Zn)) + geom_histogram(bins=50)

multiplot(p1, p2, p3, p4, cols=2)
```

## "shapes" of "distributions"

To use a histogram, *glance* at it and look for any of the following (without getting fooled by plot artefacts):

```{r}
# These plots use base R plots, which I tend to avoid
layout(matrix(1:4, 2, 2, byrow=TRUE))
plot(density(rnorm(10000), bw=1), axes=FALSE, xlab="", ylab="", main=""); box(); title("Symmetric")
plot(density(rweibull(10000, 1.5, 10), bw=2), axes=FALSE, xlab="", ylab="", main=""); box(); title("Right skewed")
plot(density(-rweibull(10000, 2, 10), bw=2), axes=FALSE, xlab="", ylab="", main=""); box(); title("Left skewed")
plot(density(c(rnorm(10000, 0, 1), rnorm(10000, 7, 2)), bw=1), axes=FALSE, xlab="", ylab="", main=""); box(); title("Multimodal")
```

## transforming variables

Apply log or square root to a variable will change the shape of the empirical distribution, e.g. transform right-skewed to symmetric. 

```{r}
oil %>% 
  filter(Fe < 100) %>% 
  ggplot(aes(x=log(Fe+1))) + geom_histogram(bins = 35)
```

## boxplots

A special plot of these (or similar) five numbers:
$$\min \qquad 25^{th}\text{ percentile}\qquad \text{ median} \qquad 75^{th}\text{ percentile} \qquad\max$$
is called a *boxplot*. Often the extreme values are shown individually (see documentation for the (irrelevant) details.)

Best as *side-by-side* boxplots with more than one varaible on the same scale.

## boxplot example - I

```{r, message=FALSE}
oil_long %>% 
  ggplot(aes(x=element, y=ppm)) + geom_boxplot()
```

## boxplot example - II 

```{r, message=FALSE}
oil_long %>% 
  filter(!element %in% c("Ca", "Ph", "Zn")) %>%
  ggplot(aes(x=element, y=ppm)) + geom_boxplot()
```

## scatterplot

A graphic for two numerical variables, e.g. `Fe` and `Si`

```{r}
oil %>% 
  ggplot(aes(x=Fe, y=Si)) + geom_point()
```

## `Fe` vs `Si` without the "outliers"

```{r}
oil %>% 
  filter(Fe < 100) %>% 
  ggplot(aes(x=Fe, y=Si)) + geom_point()
```

## alternatively, on a log-log scale

```{r}
oil %>% 
  ggplot(aes(x=log(Fe+1), y=log(Si+1))) + geom_point()
```

## "small multiples" through faceting

A powerful exploratory tool is to make a grid of small plots on subsets of the data.

```{r}
oil %>% 
  ggplot(aes(x=log(Fe+1), y=log(Si+1))) +
  facet_wrap(~Ident) + 
  geom_point()
```

## what about that "Date" variable...(!)

```{r}
oil %>% 
  ggplot(aes(x=Date, y=Zn)) + geom_point()
```

## `Fe` versus `Date`, facet by `Ident`

```{r}
oil %>% 
  ggplot(aes(x=Date, y=Fe)) + facet_wrap(~Ident) + geom_point()
```